---
title: "State Space Models"
date: "2025-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let $X_t$ be the **true** concentration of chlorophyll-a on day $t$ and $Y_t$ be the **measured** value on day $t$.

These are all our covariates:

  - $Z_{DO, t}$ = Daily mean dissolved oxygen on day $t$
  - $Z_{pH, t}$ = Daily mean pH
  - $Z_{\text{turb}, t}$ = Daily mean turbidity
  - $Z_{\text{temp}, t}$ = Daily mean water temperature
  - $Z_{lr, t}$ = Daily mean longwave radiation
  - $Z_{sr, t}$ = Daily mean shortwave radiation
  - $Z_{prec, t}$ = Daily total precipitation flux
  
# Set up

```{r}
library(rjags)
library(ecoforecastR)

source("01_download_data.R")
source("02_combine_data.R")
```

# Null Model(s)

## Random Walk Model

Data model:
$$ Y_t \sim N(X_t, \tau_\text{obs}) $$

Process model:
$$ X_{t+1} \sim N(X_t, \tau_\text{add}) $$

Priors:
$$ X_1 \sim N(X_{ic}, \tau_{ic}) $$
$$ \tau_\text{obs} \sim \text{Gamma}(a_\text{obs}, r_\text{obs}) $$
$$ \tau_\text{add} \sim \text{Gamma}(a_\text{add}, r_\text{add}) $$


## Previous Year's Chorophyll-a Model


Data model:
$$ Y_t \sim N(X_t, \tau_\text{obs}) $$

Process model:
$$ X_{t} \sim N(X_{t-365}, \tau_\text{add}) $$

Priors:
$$ X_1 \sim N(X_{ic}, \tau_{ic}) $$
$$ \tau_\text{obs} \sim \text{Gamma}(a_\text{obs}, r_\text{obs}) $$
$$ \tau_\text{add} \sim \text{Gamma}(a_\text{add}, r_\text{add}) $$


# Internal Factors Model


Data model:
$$ Y_t \sim N(X_t, \tau_\text{obs}) $$

Process model:
$$ X_{t+1} \sim N(X_t + \beta_0 + \beta_{DO} Z_{DO, t} + \beta_{pH}Z_{pH, t} + \beta_{\text{turb}}Z_{\text{turb}, t} + \beta_X X_t, \tau_\text{add}) $$

Priors:
$$ X_1 \sim N(X_{ic}, \tau_{ic}) $$
$$ \tau_\text{obs} \sim \text{Gamma}(a_\text{obs}, r_\text{obs}) $$
$$ \tau_\text{add} \sim \text{Gamma}(a_\text{add}, r_\text{add}) $$

# External Factors Model

Data model:
$$ Y_t \sim N(X_t, \tau_\text{obs}) $$

Process model:
$$ X_{t+1} \sim N(\beta_0 + \beta_{\text{temp}} Z_{\text{temp}, t} + \beta_{lr}Z_{lr, t} + \beta_{sr}Z_{sr, t} + \beta_{\text{prec}} Z_{\text{prec}, t}, \tau_\text{add}) $$

Priors:
$$ X_1 \sim N(X_{ic}, \tau_{ic}) $$
$$ \tau_\text{obs} \sim \text{Gamma}(a_\text{obs}, r_\text{obs}) $$
$$ \tau_\text{add} \sim \text{Gamma}(a_\text{add}, r_\text{add}) $$

# Combined Factors Model

Data model:
$$ Y_t \sim N(X_t, \tau_\text{obs}) $$

Process model:
$$ X_{t+1} \sim N(X_t + \beta_0 + \beta_{\text{temp}} Z_{\text{temp}, t} + \beta_{\text{prec}} Z_{\text{prec}, t} + \beta_X X_t, \tau_\text{add}) $$

Priors:
$$ X_1 \sim N(X_{ic}, \tau_{ic}) $$
$$ \tau_\text{obs} \sim \text{Gamma}(a_\text{obs}, r_\text{obs}) $$
$$ \tau_\text{add} \sim \text{Gamma}(a_\text{add}, r_\text{add}) $$


## Fitting the Model

```{r}
# set up data object for ecoforecastR
data <- list(y = cleaned_data$chla,
             n = length(cleaned_data$chla),      ## data
             temp = cleaned_data$temperature,
             longrad = cleaned_data$surface_downwelling_longwave_flux_in_air,
             shortrad = cleaned_data$surface_downwelling_shortwave_flux_in_air,
             precip = cleaned_data$precipitation_flux,
             x_ic=1,tau_ic=100,      ## initial condition prior
             a_obs=1,r_obs=1,           ## obs error prior
             a_add=1,r_add=1            ## process error prior
)

## fit the model
model2 <- "~ 1 + X + temp + precip"
ef.out.combined <- ecoforecastR::fit_dlm(model=list(obs="y",fixed=model2),data)

#save(ef.out.combined, file = "combined_factors.RData")
```

## Parameter Diagnostics

```{r, fig.height=10, fig.width=10}
#load("combined_factors.RData")

## parameter diagnostics
params <- window(ef.out.combined$params,start=1000) ## remove burn-in
plot(params)
```

```{r}
summary(params)
cor(as.matrix(params))
pairs(as.matrix(params))
```

## Time-Series Plot

```{r}
## confidence interval
time.rng = c(1,nrow(cleaned_data))       ## you can adjust this line to zoom in and out on specific time intervals

out <- as.matrix(ef.out.combined$predict)
ci <- apply(out,2,quantile,c(0.025,0.5,0.975))

plot(cleaned_data$datetime,ci[2,],type='n',ylim=range(cleaned_data$chla,na.rm=TRUE),
     #log='y',
     xlim=cleaned_data$datetime[time.rng],
     xlab = "Time",
     ylab = "Chlorophyll-a",
     main = "Combined Factors Model")
## adjust x-axis label to be monthly if zoomed
if(diff(time.rng) < 100){ 
  axis.Date(1, at=seq(cleaned_data$datetime[time.rng[1]],cleaned_data$datetime[time.rng[2]],by='month'), format = "%Y-%m")
}
ecoforecastR::ciEnvelope(cleaned_data$datetime,ci[1,],ci[3,],col=ecoforecastR::col.alpha("lightBlue",0.75))
points(cleaned_data$datetime, cleaned_data$chla,pch="+",cex=0.5)
```

```{r}
# Set up data object (with NA handling built-in)
data_ext <- list(
  y = cleaned_data$chla,
  n = length(cleaned_data$chla),
  temp = cleaned_data$temperature,
  longrad = cleaned_data$surface_downwelling_longwave_flux_in_air,
  shortrad = cleaned_data$surface_downwelling_shortwave_flux_in_air,
  precip = cleaned_data$precipitation_flux,
  x_ic = 1, tau_ic = 100,
  a_obs = 1, r_obs = 1,
  a_add = 1, r_add = 1
)

# Fit the model — this version has no X term
model_ext <- "~ 1 + temp + longrad + shortrad + precip"

ef.out.external <- ecoforecastR::fit_dlm(
  model = list(obs = "y", fixed = model_ext),
  data = data_ext
)

# Optional: save results
# save(ef.out.external, file = "external_factors.RData")
```


```{r}
# Discard burn-in
params_external <- window(ef.out.external$params, start = 1000)

# Plot and summarize
plot(params_external)
summary(params_external)
cor(as.matrix(params_external))
```

```{r}
## confidence interval
time.rng = c(1, nrow(cleaned_data))  ## you can adjust this line to zoom in and out on specific time intervals

# Get posterior predictions from the external factors model
out_ext <- as.matrix(ef.out.external$predict)
ci_ext <- apply(out_ext, 2, quantile, c(0.025, 0.5, 0.975))

# Plot the time series with confidence intervals
plot(cleaned_data$datetime, ci_ext[2,], type='n', ylim=range(cleaned_data$chla, na.rm=TRUE),
     #log='y',
     xlim=cleaned_data$datetime[time.rng],
     xlab = "Time",
     ylab = "Chlorophyll-a",
     main = "External Factors Model")

# Adjust x-axis label to be monthly if zoomed
if(diff(time.rng) < 100){ 
  axis.Date(1, at=seq(cleaned_data$datetime[time.rng[1]], cleaned_data$datetime[time.rng[2]], by='month'), format = "%Y-%m")
}

# Add the confidence envelope
ecoforecastR::ciEnvelope(cleaned_data$datetime, ci_ext[1,], ci_ext[3,], col=ecoforecastR::col.alpha("lightBlue", 0.75))

# Plot the original data points
points(cleaned_data$datetime, cleaned_data$chla, pch="+", cex=0.5)
```
