---
title: "Ensemble Forecast and Uncertainty Analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--- NOTE: Run `Main.R` in order to generate the files needed to knit this document --->


# Setup

```{r}
library(tidyverse)
library(ecoforecastR)
library(rjags)
library(lubridate)

load("cleaned_data.RData")
load("combined_factors_noburnin.RData") #loads in params and predict
load("precip_ens_forecast.RData") # loads in esemble precipitation forecast
```


# Forecast Start Date

Insert discussion TODO

```{r}
NT = 30 # forecast length
hist_dates <- as_date(cleaned_data$datetime)
forecast_dates <- seq.Date(from = tail(hist_dates, 1) + 1, by = "day", length.out = NT)
dates <- c(hist_dates, forecast_dates)
```

# Driver Forecasts

For precipitaiton flux, we are using a 30 member ensemble forecast from NOAA GEFS. The ensemble and their mean are plotted below. 

```{r}
# Filter down to just the 30 days we are forecasting for
precip_ens_forecast <- precip_ens_forecast %>%
  filter(date %in% forecast_dates)

plot(precip_ens_forecast$date,
     precip_ens_forecast$ens_0,
     type = 'l', col = "gray", ylim = range(precip_ens_forecast[ , grepl("ens_", names(precip_ens_forecast))], na.rm = TRUE),
     xlab = "Date", ylab = "Precipitation Forecast", main = "NOAA Ensemble Forecast with Mean")

for(i in 0:30) {
  col_name <- paste0("ens_", i)
  lines(precip_ens_forecast$date,
        precip_ens_forecast[[col_name]],
        col = "gray")
}

lines(precip_ens_forecast$date, precip_ens_forecast$ens_mean, col = "blue", lwd = 3)
```


For water temperature, we will be using historical data for the dates being forecasted to create an ensemble forecast. 

```{r}


target_start <- forecast_dates[1]  #only month/day matter

# Step 1: Prepare the data with date components
temp_with_dates <- cleaned_data %>%
  mutate(date = as_date(datetime),
         year = year(date),
         doy = yday(date))

# Step 2: Loop over years and build the 30-day slices
years <- unique(temp_with_dates$year)

# Helper to extract 30-day windows for each year
get_30_day_window <- function(y) {
  start_date <- ymd(paste0(y, "-", month(target_start), "-", day(target_start)))
  end_date <- start_date + days(NT - 1)
  
  temp_with_dates %>%
    filter(date >= start_date & date <= end_date & year == y) %>%
    arrange(date) %>%
    mutate(day_index = row_number()) %>%
    select(year, day_index, temperature)
}

# Step 3: Combine all years
all_years_30d <- bind_rows(lapply(years, get_30_day_window))

# Step 4: Pivot to wide format, one column per year
temp_ensemble_df <- all_years_30d %>%
  pivot_wider(names_from = year, values_from = temperature, names_prefix = "yr_")


temp_ensemble_df$date <- forecast_dates

# If any years have no data at all, remove them
temp_ensemble_df <- temp_ensemble_df %>%
    select(where(~ !all(is.na(.))))

# Add a column with the mean forecast
temp_ensemble_df <- temp_ensemble_df %>%
  mutate(ens_mean = rowMeans(select(., starts_with("yr_")), na.rm = TRUE))

```


```{r}
ensemble_cols <- grep("^yr_", names(temp_ensemble_df), value = TRUE)

# Plot the first ensemble member to set up the plot
plot(temp_ensemble_df$date, temp_ensemble_df$ens_mean,
     type = "n", col = "grey", ylim = range(temp_ensemble_df[ ,ensemble_cols], na.rm = TRUE),
     xlab = "Date", ylab = "Water Temperature", main = "30-day Historical Ensemble")

# Loop through all ensemble member columns
for (col_name in ensemble_cols) {
  lines(temp_ensemble_df$date, temp_ensemble_df[[col_name]], col = "grey")
}

# Add the mean in blue
lines(temp_ensemble_df$date, temp_ensemble_df$ens_mean, col = "blue", lwd = 2)
```



# Run forecast

## Set up

```{r}
forecastN <- function(IC,
                      Nmc,
                      Betaintercept,
                      Betatemp,
                      Betaprecip,
                      BetaX,
                      temperature,
                      precipitation,
                      sd,
                      NT = 30){
  N <- matrix(NA,Nmc,NT)  ## storage
  Nprev <- IC           ## initialize
  
  for(tt in 1:NT){
    ## calculate mean
    mu = Nprev + Betaintercept + Betatemp*temperature[tt] + Betaprecip*precipitation[tt] + BetaX*Nprev   
    N[,tt] <- rnorm(Nmc,mu, sd)                         ## predict next step
    Nprev <- N[,tt]                                  ## update IC
  }
  return(N)
}
```

```{r}
Nmc = 1000         ## set number of Monte Carlo draws
predict_matrix <- as.matrix(predict)

#sample rows from mcmc chain to run foreward in time
prow = sample.int(nrow(as.matrix(params)), Nmc, replace = TRUE)

## Set up parameters
param_matrix <- as.matrix(params)
param.mean <- apply(param_matrix,2,mean) # Find mean parameter values 
```


## Initial Condition Uncertainty

```{r}
# Sample initial conditions
IC <- predict_matrix[prow, ncol(predict_matrix)]

N.ic <- forecastN(IC,
                Nmc,
                param.mean["betaIntercept"],
                param.mean["betatemp"],
                param.mean["betaprecip"],
                param.mean["betaX"],
                temp_ensemble_df$ens_mean,
                precip_ens_forecast$ens_mean,
                0)

```

## Parameter Uncertainty

```{r}
# Sample parameter values
Betatemp.sample <- param_matrix[prow, "betatemp"]
Betaprecip.sample <- param_matrix[prow, "betaprecip"]
BetaX.sample <- param_matrix[prow, "betaX"]
Betaintercept.sample <- param_matrix[prow, "betaIntercept"]


N.ic.p <- forecastN(IC,
                Nmc,
                Betaintercept.sample,
                Betatemp.sample,
                Betaprecip.sample,
                BetaX.sample,
                temp_ensemble_df$ens_mean,
                precip_ens_forecast$ens_mean,
                0)
```

## Driver Uncertainty

```{r}
## Sample driver columns with replacement

# Turn temperature forecast ensemble tibble into a matrix
temp_ens_matrix <- temp_ensemble_df %>%
  select(starts_with("yr_")) %>%
  as.matrix()
temp_cols <- sample.int(ncol(temp_ens_matrix),Nmc,replace=TRUE) # sample

# Turn precipitation forecast ensemble tibble into a matrix
precip_ens_matrix <- precip_ens_forecast %>%
  select(starts_with("ens_") & !ends_with("mean")) %>%
  as.matrix()
precip_cols <- sample.int(ncol(precip_ens_matrix),Nmc,replace=TRUE)



N.ic.p.d <- forecastN(IC,
                Nmc,
                Betaintercept.sample,
                Betatemp.sample,
                Betaprecip.sample,
                BetaX.sample,
                temp_ens_matrix[ , temp_cols],
                precip_ens_matrix[ , precip_cols],
                0)
```




## Process Uncertainty


```{r}
tau_add.sample <- param_matrix[prow, "tau_add"]
sd.sample <- 1/sqrt(tau_add.sample)

N.ic.p.d.pr <- forecastN(IC,
                Nmc,
                Betaintercept.sample,
                Betatemp.sample,
                Betaprecip.sample,
                BetaX.sample,
                temp_ens_matrix[ , temp_cols],
                precip_ens_matrix[ , precip_cols],
                sd.sample)
```

# Plot

```{r}
## you can adjust this line to zoom in and out on specific time intervals
time.rng = c(2600,nrow(cleaned_data)+30)  

## You can adjust this line to adjust the y-axis zoom
y.rng <- range(cleaned_data$chla,na.rm=TRUE)
#y.rng <- c(0,5)


## Plot historical fit data
out <- predict_matrix
ci <- apply(out,2,quantile,c(0.025,0.5,0.975))

# Time interval set up
dates <-  as.Date(cleaned_data$datetime)
dates <- c(dates, seq.Date(from = tail(dates, 1) + 1, by = "day", length.out = 30))
```


```{r}
plot(dates, 1:length(dates) ,type='n',
     ylim = y.rng,
     #log='y',
     xlim=dates[time.rng],
     xlab = "Time",
     ylab = "Chlorophyll-a",
     main = "Combined Factors Model")
## adjust x-axis label to be monthly if zoomed
#if(diff(time.rng) < 100){ 
#  axis.Date(1, at=seq(cleaned_data$datetime[time.rng[1]],cleaned_data$datetime[time.rng[2]],by='month'), format = "%Y-%m")
#}
ecoforecastR::ciEnvelope(as.Date(cleaned_data$datetime),ci[1,],ci[3,],col=ecoforecastR::col.alpha("lightBlue",0.75))
points(as.Date(cleaned_data$datetime), cleaned_data$chla,pch="+",cex=0.5)




## Add forecast and uncertainties
N.ic.ci = apply(N.ic,2,quantile,c(0.025,0.5,0.975))
N.ic.p.ci = apply(N.ic.p,2,quantile,c(0.025,0.5,0.975))
N.ic.p.d.ci = apply(N.ic.p.d,2,quantile,c(0.025,0.5,0.975))
N.ic.p.d.pr.ci = apply(N.ic.p.d.pr,2,quantile,c(0.025,0.5,0.975))

ecoforecastR::ciEnvelope(tail(dates, 30),N.ic.p.d.pr.ci[1,],N.ic.p.d.pr.ci[3,],col= "red")
ecoforecastR::ciEnvelope(tail(dates, 30),N.ic.p.d.ci[1,],N.ic.p.d.ci[3,],col= "green")
ecoforecastR::ciEnvelope(tail(dates, 30),N.ic.p.ci[1,],N.ic.p.ci[3,],col= "blue")
ecoforecastR::ciEnvelope(tail(dates, 30),N.ic.ci[1,],N.ic.ci[3,],col= "orange")


lines(tail(dates, 30),N.ic.ci[2,],lwd=0.5)

legend("topleft", legend = c("IC Uncertainty", "Parameter Uncertainty", "Driver Uncertainty", "Process Uncertainty"), fill =  c("orange", "blue", "green", "red"))
```